---
title: "Final Project"
format: revealjs
editor: visual
author: "Karan Chaudhari"
---
## What is Visual Statistical Learning?
### Definition

-   Visual Statistical Learning (Visual SL) is the cognitive ability to implicitly extract statistical patterns and regularities from visual stimuli in our environment.

### Why It Matters

-   It underpins how we make sense of the world, allowing us to predict future events based on past sequences without conscious effort. It is a fundamental component of cognitive development.

## Method/Data Source

-   To examine Visual SL in young children using both online measures (reaction times) and offline measures (recognition tasks). 
-   The study investigates whether this learning is purely implicit or involves active engagement. It does this in 2 experiments, of which we only look at the first one.
-   The first experiment was a visual experiment, where they had kids press a button if they saw a red alien enter a spaceship.
-   They weren't told that the aliens were actually in sets of 3. (Ex. 2 blue aliens then a red alien)

## Method/Data Source

-   The researchers were looking into how fast the children learned the patterns of the triplets, and they used the reaction times of the children for that purpose.
-   Original paper analyzed overall Visual SL differences, but I want to examine how quickly the learning effect emerges with respect to the ages of the children.

## Motivating question(s)

-   Does the rate at which children reduce their reaction times to predictable stimuli, relative to unpredictable stimuli, change as a function of age/gender?

## Hypothesis

-   Age difference: Older children, due to superior attention control and processing speed, will exhibit a steeper (faster) reduction in the predictable-unpredictable RT difference across trials, reflecting more efficient, active engagement with the implicit task structure.

-   Gender difference:  While I don't predict there being any differences between genders, I am curious about whether gender acts as a moderating factor in the rate of behavioral improvement.

## Learning Rate vs. Age in children 3-9 (Experiemnt 1)

The learning rates of the children are calculated by taking the average of their change in reaction time over the course of the first experiment.

```{r, include=FALSE}
library(tidyverse)
```
```{r, echo=FALSE}
df_processed <- read_delim("VSL_main_data.csv", delim = ";") |> 
  mutate(
    diff1 = D1 - I1,
    diff2 = D2 - I2,
    diff3 = D3 - I3,
    diff4 = D4 - I4
  ) |> 
  select(ID, age, starts_with("diff"), sex) |> 
  pivot_longer(
    cols = starts_with("diff"),
    names_to = "block_str",
    values_to = "rt_diff"
  ) |> 
  mutate(
    block = as.numeric(str_extract(block_str, "\\d"))
  )
subject_slopes <- df_processed |> 
  group_by(ID, age) |> 
  summarise(
    learning_rate = coef(lm(rt_diff ~ block))[2],
    .groups = "drop"
  ) |> 
  filter(!is.na(learning_rate))
ggplot(subject_slopes, aes(x = age, y = learning_rate)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Learning Rate vs. Age",
    subtitle = "Rate at which children increase the gap between Unpredictable and Predictable RTs",
    x = "Age (months)",
    y = "Learning Rate (Slope of RT Difference)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )
```

## Observations

-   The learning rate does not significantly change as a function of age.

-   This likely is mirrored over in the gender category as well.


## Learning rate vs. Gender (Experment 1)

Learning rates are calculated same as before

```{r, echo=FALSE}
subject_slopes <- df_processed |> 
  group_by(ID, sex) |> 
  summarise(
    learning_rate = coef(lm(rt_diff ~ block))[2],
    .groups = "drop"
  ) |> 
  filter(!is.na(learning_rate))
ggplot(subject_slopes, aes(x = sex, y = learning_rate)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Learning Rate vs. Sex",
    subtitle = "Rate at which children increase the gap between Unpredictable and Predictable RTs",
    x = "Age (months)",
    y = "Learning Rate (Slope of RT Difference)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )
```

## Observations

- As expected, gender made no difference in the average learning rate of the children. However, there is a difference between the extremes of the genders, with girls having smaller extremes than boys. 

## Conclusion

- Based on the statistical visualization of the data, the answer to the research question is no. The learning rate of children does not appear to change significantly as a function of age or sex.

- In both graphs, the wide dispersion of blue dots suggests that "Learning Rate" is likely driven by individual differences rather than these specific demographic factors.